Conservator's Corner

### 2019-01-13 pnsac/newsletter/article_EditorsNotes.tex
January 2019 changes and additions.

As the President has noted the Association has, through its volunteers,
expended thousands of hours on the North Star's restoration since 2004. In the
last couple of years, however, various developments at CASM have resulted in a
diminution in the amount of restoration work we have been able to do. The result
has been a decline in the number of volunteers working on the aircraft, and a
corresponding decline in our membership. This in turn has led to an overall
decline in the Association's activities, including the publication of this
newsletter. CASM has made some major changes to its personnel, including those
directly responsible for the restoration of the North Star. The Association is
already experiencing a positive change in its relationship with the Museum which
can only be to the good. We look forward to a renewed partnership with CASM to
our mutual benefit.

%\vspace{12pt}

>>>>

In this issue we are recognizing the 70th anniversary of a very special event
in Canadian aviation history and one which involved a North Star aircraft. The
first non-stop transcontinental flight took place on January 15, 1949 and
readers will find an article dealing with that event in this issue. In
addition, your Association, in partnership with the Canadian Aviation and Space
Museum will be holding a members meeting, and a public viewing of the North
Star on January 19, 2019 to mark this event.

We are also recognizing another one of our long standing members in this
issue. Ted Devey has been active in our organization since the very early
days. We are indeed fortunate that we have a number of other members who have
been with us since the beginning. In 2003 the Ottawa Citizen predicted "a
bright future" for our nascent organization. A copy of the Citizen article is
included in this issue. Not surprisingly, given the challenges involved in
making this project work, the "future" did not always appear as "bright" as
Polaris. However, our ever optimistic President reports that 2018 was
characterized by positive developments in the Project, and our Association's
relationship with CASM.

The Museum's conservator, in the "Conservator's Corner" has contributed a
thoughtful article based on the classic 70's book, \textit{Zen and the Art of
Motorcycle Maintenance} which I am sure
that our readers will enjoy. Good
reading.

%Our Association was incorporated in 2003. We are fortunate that a
%number of the original members of the organization, such as Tim
%Timmins, are still involved as they can provide an historical
%perspective. As Tim says, in the article "Our Members," there have been
%many bumps in the road, and the fact that the aircraft is still a long
%way from completion after 16 years speaks for itself. In January 2017,
%R\'{e}jean (Rej) Demers took up his position as CASM's Conservation Project
%Manager. Rej's first contribution to the Chronicle can be found in the
%"Conservator's Corner." The restoration of the North Star is a priority
%for Rej. He has spent the first year of his mandate gathering
%information and developing systems and procedures which will, in the
%long run, move the restoration of the aircraft forward in a coherent
%manner. I have previously stated the obvious and that is the health and
%vigour of the Association is dependent on the restoration status of the
%aircraft. In the last few years there have been as many bumps in the
%road as there are potholes in Ottawa in the spring. The President is
%very optimistic we have reached a real turning point (see the
%"President's Notes") and wouldn't that be great.

### 2019-01-13 pnsac/newsletter/article_Feature1.tex
January 2019 changes and additions.

\title{Reunion of Giants}
%\subtitle{Part 4}
\author{Bruce Grant}

@@

% \textit{Photos courtesy of Library Archives Canada and 426 Thunderbird
%  Sqn, RCAF.}\\

The "Reunion of Giants" video came to the Canadian Aviation and Space Museum's
theatre on May 14th, courtesy of the Canadian Warplane Heritage Museum in
Hamilton. CWHM sent along the pilot Captain Leon Evans (retired Air Canada), and
Dave Rohrer, CEO, as well as Morgan Elliott the videographer. Also present were
two World War Two decorated Lancaster pilots Don McKechnie and Bill Button. With
almost a full house of 200 it was a very successful Project North Star public
event.

The outstanding feature of the story is the audacity of CWHM to fly Canada's
only airworthy Lancaster to England for a reunion with the only other flying
Lancaster in the world. The Lanc is the flagship of the CWHM, its biggest
attraction and an irreplaceable historic artifact. In a typical year it flies
only about fifty hours, attending aviation events in Canada and every move is
calculated some years in advance.

In England, the arrival over the North Atlantic route of our Lanc was embraced
as a tribute to the quality of the machine and the courage of its crew. It
reinforced the link with Canada which provided so much support to Great Britain
during the war.

While on a final approach in England, the supercharger on engine \#3 went to
pieces and the plane landed with a trail of smoke. A replacement engine would
have to be found. The replacement that turned up was a Rolls Royce Merlin that
would have to be substituted for the Packard Merlin in our Canadian-built
Lancaster. This presented a challenge in sorting out the elements of QEC with
incompatible Whitworth and SAE fasteners. The new engine would be a "bitsa"
(bitsa this and bitsa that). They did it in about three days and got the plane
back in the air.

On the return flight, stronger than anticipated headwinds raised the risk of
running out of gas over the Davis Strait. They diverted to Narsarsuaq in
Greenland.

Following the showing of the film there was a spirited question and answer
session with CWHM's representatives moderately by Mike Pearson (retired Air
Canada pilot of the Gimli glider fame). There is no doubt that all those in
attendance enjoyed themselves thoroughly and gained some understanding of what
it meant for those involved in this project.

At Project North Star, we have some appreciation idea of the scale of CWHM's
accomplishment. As our members know we are in the twelfth year of our project to
restore the last surviving North Star. But, we are restoring it only as a museum
artifact, not an airworthy machine. The video is available for sale at
www.warplane.com and the trailer can be viewed ath www.suddenlyseemore.com .

@@

%\textit{Reprinted from the March/April 2014 edition of "Legion Magazine," with the kind permission of Jennifer Morse, General Manager of Canvet Publications, and Tom MacGregor.}

>>>>

\title{Volunteers Assure a Bright Future for Classic Plane}
%\subtitle{Part 4}
\author{Randall Denley}

@@

\end{multicols}

\begin{figure*}[htbp]
	\vspace{2em}
	\centering
	%name of the graphic, without the path AND in EPS format:
	\includegraphics[scale=0.8]{holmgren-timmin-scaled.png}
	%caption of the figure 
	\caption*{\small \em TIm Timmins, left, and Robert Holmgren are forming a
group of skilled volunteers to restore a 55-year-old North Star, the first
Canadian-made plane capable of transcontinental flight.}
	%label of the figure, which has to correspond to \ref{}:
	\label{fig:timmins-holmgren}
\end{figure*}

\begin{multicols}{2}

\textit{This article was published in the Ottawa Citizen on February 1, 2003
under the byline of Randall Denley. It was brought to our attention recently by
Drew Hodge. The article is republished with the express permission of the
Ottawa Citizen, a division of Postmedia Network Inc. (ed.).
}\\

The world's last classic Canadair North Star sits, forlorn, on the tarmac at
the Canada Aviation Museum. It's been parked there since 1966, along with half
a dozen other orphans of the museum, planes that it has no money to restore and
that are too big to bring inside the crowded building.

The paint on the North Star is faded and peeling.Some gaps in the wings are
closed with wire to keep out the birds that have made it home for years.The
engines are protected with pieces of plywood. The interior requires a complete
renovation.

It's a shame to see an important part of Canada's aviation history slide into
this kind of ruin, Thanks to the efforts of two Ottawa men, however, the North
Star finally has a bright future.

Robert Holmgren and Tim Timmins are forming a group of skilled volunteers to
restore the North Star. This kind of volunteer effort will be a first for the
museum, says director general Anthony Smyth, who is supporting the effort,
along with aviation companies including Air Canada and Boeing.

The museum is showing commendable flexibility because the traditional view in
the museum world is that restoration is work that can only be done by trained
experts. Unfortunately, the museum can't afford that kind of work, and that's
one of the reasons the plane has been sitting there for more than 35 years.

This is no small undertaking. The 55-year-old North Star 1-ST is a sizeable
plane, nearly 30 metres long and with a wing span of 35 metres. Restoring it
will be a yearlong process that would cost \$1 million if it were done by paid
staff.

The North Star is an important plane in Canada's aviation history, the first
Canadian-,made plane capable of transcontinental flight. It's considered the
plane that launched the postwar Canadian aviation industry; 70 of them were
built. Brought into service after the Second World War, North Stars carried
passengers for several airlines and troops for the Canadian military.  North
Stars served to airlift Canadian troops to Korea and provided transportation
for VIPs visiting Canada. The plane at the museum belonged to the RCAF.

Holmgren is a retired Air Canada maintenance expert, and Timmins is a former
RCAF navigator who flew on the North Stars. Timmins's former squadron had taken
an interest in the North Star, and Holmgren, who was giving his time as a
volunteer at the museum, pushed it forward.

"We'd like to see it brought back to something like its former self," Holmgren
says.

A temporary structure, perhaps a bubble like the one over the playing field at
Lansdowne Park, will be necessary before the restoration can begin. The plane
is too large to fit inside the museum's existing building and also too big to
be easily transported elsewhere.

Holmgren and Timmins need about 200 volunteers to make the project a success.
Holmgren has approached Air Canada, Bombardier and Boeing to get the help of
past and present employees. About 40 volunteers have already come forward.

The key need is skilled craftsmen with up-to-date large aircraft experience and
who live in or near Ottawa.

Some people who can come in almost every day to run the project are required. 
There is also a demand for people to do record-keeping and research.

The group is hoping to attract national attention for the project, both to
provide volunteer expertise and to help raise money.  This is a plan to help a
national museum, after all.  Smyth says the museum will be involved with a
fundraising campaign, but it's early yet to say how much is required. It
depends partly on how much time and material are donated by corporations. A
full-time salaried project manager will be hired by the museum.

The project will take a conservation approach, retaining as much of the
original material as possible, Smyth says. The restored plane will be
structurally complete and faithful to the original. Returning the plane to
flyable condition is too expensive to be feasible and the museum wouldn't want
to take a chance flying the last remaining version of such an important plane.

Restoring the North Star is just the beginning of what Holmgren and his
associates hope to accomplish. The museum has another half dozen planes outside
that require restoration. Restoring them all will take nine or ten years. The
Museum's new addition, which it hopes to have open by December of this year,
will be large enough to finally bring the North Star indoors.

%If you want to help with the project, contact Robert Holmgren at 748-5972 or
%robertholmgren@rogers.com. The group also has a Web site,
%www.projectnorthstar.ca.

%\address{Web site:
%{\normalfont\color{blue}\texttt{\url{http://www.projectnorthstar.ca}}}\\
%	General enquiries:\email{info@projectnorthstar.ca}}
%

@@

### 2019-01-13 pnsac/newsletter/article_MemberInterview.tex
January 2019 changes and additions.

\title{Our Members}
\author{Interview with Austin J Timmins}

@@

My aviation experience includes 36 years service in the RCAF and nine
years with the Canadian Aeronautics and Space Institute. In the RCAF I
completed radio navigator training, which included communications,
navigation and air gunner. I later qualified as an airborne
interception navigator. My operational experience includes air
transport, search and rescue and training. My assignments included
Navigation Leader, Aircrew Leader, Squadron Commander, Air Command
Navigator and Deputy Commander (COO) of the Air Transport Group. I
logged 9500 hours, mostly on the following aircraft: Beechcraft,
Dakota, North Star, Mitchell, Yukon, Hercules, and Boeing 707. My
staff assignments in Europe included movement control (traffic
management) for 1 Air Division and a tour at The NATO Defense College
in Rome.

On retirement from the RCAF I participated in a Transport
Canada A Base review of several airports. I then secured full
employment with Canadian Aeronautics and Space Institute as Executive
Director, retiring in 1995.

@@

My involvement with efforts to conserve North Star 17515 predates
Project North Star. Ottawa area members of the 426 Thunderbird Squadron
Association, mostly retired North Star aircrew, met with the National
Aviation Museum staff to express their concern about the state of the
North Star and proposed volunteer involvement in a preservation effort.
These expressions of concern and interest led to a meeting between the
Association representatives, where General Adamson presented a proposal
for the restoration of the aircraft. It was rejected but assurance was
given that the North Star would be first priority when the new hangar
was completed Also a North Star Trust Fund would be established. Robert
Holmgren, an observer at the meeting, then proceeded to develop a
proposal for volunteers to restore the North Star.

@@

I have been involved from the beginning of efforts to establish PNS,
first as a member of a Steering Committee, led by Robert Holmgren,
later, Vice President and then President of PNSAC. My contacts in
aviation provided much needed advice and support, including access to
initial funding for PNSAC.

@@

The highlight of my involvement has to be the establishment of PNS and
PNSAC.

@@

PNS is now in its 16th year. It has been a work in progress with many
bumps along the way. I spent a lot of time and effort, along with other
members of the Executive, evening out these bumps. Establishing and
maintaining effective channels of communication with CASM and CSTM
required constant attention. Much effort was directed to fund raising,
publicity, and, finding volunteers. In the beginning, every date
brought a new challenge.

>>>>

\title{Our Members}
\author{Interview with Ted Devey}

@@

My service background was naval. Professionally I was an electrical engineer
with the Department of Communications specializing in radio spectrum
engineering. After retirement in 1989 I sought activities that interested me
such as helping to install a pipe organ in a theatre in Renfrew, working on
restoring a sailing yacht, which unfortunately stopped due to lack of money. I
also took courses in welding, machine shop and woodworking at Algonquin College
to enhance my manual skills.

@@

I attended a public meeting in the Museum in 2003 in which it was indicated
that efforts were being made to restore the Museum's North Star to a showroom
condition. In 2004 the project finally was approved and work started.
Professional work had to be carried out to remove asbestos and many dead birds
and their mess. At first I used my welding skills in adapting steps/platforms
for working on engines on the aircraft. In winter, seats were removed from the
cockpit and brought into the shop where they were dismantled, treated and
cleaned, painted and finally spray painted. I found the work interesting and
satisfying and so I was hooked into the project.

@@

In 2006 I expressed interest in the work on the four Roll-Royce Merlin engines.
As it turned out I was asked if I would take on the job of Merlin Crew Chief
and agreed with the proviso that I would turn the task over if someone came
along who was properly qualified. When Garry Dupont retired from the RCMP as
head AME at the RCMP Hangar at Uplands, I offered the position to him which he
took over. From 2006 until the summer 2018 I did only engine work on the four
engines. Unfortunately my limited mobility prevented me from helping with the
more recent work that was being done behind the firewall of \#4 engine. So I
went into the Conservation shop to start doing different work.

@@

I think that the highlight of my involvement was back in 2006. To work on the
engines we needed two engine stands, one to transfer the entire engine assembly
from the aircraft for dismantling to the core engine and a second stand to
mount the core engine and further dismantle it to bare components. This had to
be a rotary affair so that the engine could be positioned for whatever
operation was being done. The first stand had to be designed and built from
scratch. The design was an adaptation of the engine stand located next to the
Lancaster bomber on the Museum floor with ample room for a person to work
underneath. A specification was drawn up for the various pieces of steel to be
cut precisely by the steel supplier and to be delivered to the RCMP Hangar at
Uplands, as instructed by Garry Dupont. On a Saturday a group of volunteers
assembled at the RCMP Hangar and assembled the horizontal and vertical sections
of the stand with Garry doing the welding. Following this the stand was
delivered to the Museum's welding shop for final assembly, then into the main
shop where it was finally assembled and painted. Everything fit as intended.

The rotary stand was located in the storage area of the Science and Tech
Museum. The base was extended to the required length and a new frame was built
to hold the core engine. That everything fit together so well from the start
was the highlight of the project for me.

@@

The most challenging part of the job was working on engine \#1, the port outer.
Since I was not a mechanic, I had to plan the disassembly of the engine and
carry it out with the help of crew members under the strict guidance of the
then Shop Supervisor Mike Irvin. He was a hard taskmaster. This was not an
engine that had just been removed from an operating aircraft for overhaul. The
North Star and its engines had been left out in the Great Canadian outdoors for
39 years without any attempt at protection from the four seasons of weather
during those years. Corrosion was rampant. In each of the twelve cylinders the
piston rings were rusted to the cylinder liners. Many hours of soaking the
pistons with Liquid Wrench, the use of compressed air and other techniques had
to be employed to remove the pistons from the cylinders. There were other
issues too that made the overall disassembly of the engine very difficult. The
foregoing applied to each of the other three engines, so it has taken about 12
years to process the four Merlins to a preserved non-running condition. The
first engine I found the most challenging as we had to learn as we went along.
The remaining engines were very difficult but the experience gained with \#1
engine was valuable in restoring the others.

### 2019-01-13 pnsac/newsletter/article_PresReport.tex
January 2019 changes and additions.

These are exciting times to be a member of the Project North Star Association.
We  have recently taken part in the 2018 Doors Open Ottawa. This is a major
event for the Museum and PNSAC volunteers were able to make a significant
contribution to the success of the two open days. Doors Open has become second
only to Canada Day in its importance to the Museum. (more on Open Doors 2018
later in this issue).

A few days before Canada Day the aircraft was brought outside. Unfortunately,
the day itself was extremely hot and the number of visitors was well down from
previous years. Nonetheless our volunteers did a yeoman service in talking to
the public about our restoration work.

Although showing the aircraft to the public is very important to us, our main
purpose is to restore the North Star to museum condition. Much of our work is
detailed, painstaking and unspectacular such as restoring a small part of the
aircraft which can take many days or weeks. At other times making progress is
very visible. We are in such a situation now.

At the end of May we had one of our regular meetings with Chris Kitzan (Director
General of CASM) together with Neil Raynor (VP of PNSAC), Cedric St-Amour (CASM
Volunteer Coordinator) and Reg Demers (Our CASM supervisor There was a full
exchange of ideas, suggestions for bringing  the public into the Museum and ways
in which PNSAC could be involved in these initiatives and events. The meeting
ended with me having a full sheet of notes covering future planning, major
restoration work and how we can make sure we recruit younger people both as
members of PNSAC and where possible active Restoration Volunteers.

A meeting like the one above only happens when an organization is thriving, and
everybody is excited about the future. During the last 15 years PNSAC has
demonstrated that it is capable delivering very high-quality restoration work.
We have already accumulated over 80,000 hours of work and during this time we
have become a trusted partner of the Museum. Our website is now operational
again and in the next few weeks we expect to be able to update it and make it
more relevant to where we are with the project now. One of the major initiatives
taking place is to develop a diagrammatic history of the work done on the North
Star restoration and a timeline for future work on the aircraft.

In closing, I once again urge Association members to become active volunteers.
There are many volunteering opportunities both for those with major aviation
experience to those who can only volunteer with us on occasions like Canada Day
or can provide help for our many behind-the-scenes activities.

Our next Members' Meeting is scheduled for
Saturday, September 29, 2018. Do reserve this date in your calendar if you can
be in the Ottawa area on that day. The meeting will be held at the Canada
Aviation and Space Museum. More details will follow later.

>>>>

It is always interesting to look back on what was written in previous issues of
any journal. The NStar Chronicle is no exception and so before starting to
write this piece I reread what I had said during the year 2018. For once luck
was with me and as far as I can see I did not write anything that was either
untrue or not achieved during the year.

The last paragraph of my January 2018 Notes suggested that the year would see
increased activity both in the aircraft restoration and within the Association.
Both have proved correct. We have successfully recruited several new
Restoration Volunteers and our Association activities have increased
substantially.

Apart from the occasions when our members either met or have the been actively
taking part in Museum activities we have made a major step forward in that our
Facebook page has again become active under the stewardship of Chris McGuffin.
Chris is a dedicated Restoration Volunteer who can frequently be seen around
the aircraft at a considerable height complete with hardhat and safety harness.
When he returns to the ground, he often gets out his camera and takes
photographs, many of which are now appearing on Facebook. For those of our
readers who are not Facebook subscribers, I encourage you to join Facebook and
to "like" the page Project North Star. You will then see the latest updates on
our restoration progress together with the recent photographs.

Within the next few days our webmaster, Drew Hodge, will be relaunching our
website with an updated homepage and better links to other parts of the
website. Much work has gone into the redesign and we look forward to receiving
comments from users. The website is designed primarily to be a location for all
the Association's historical and archival information. We are planning to add
to the website more PDF versions of previous issues of the NStar Chronicle and
photographs which are not currently available on the site. Current information
will be sent to members by email and updates to our Facebook page.

Our first big event of 2019 will be held at the Museum on Saturday, January
19th to celebrate the 70th anniversary of the first non-stop flight across
Canada between Vancouver and Halifax. The flight was made by a North Star on
15th January 1949. Members have already been advised of the event by email and
on Facebook and I hope to see many of you joining us on that day. Although this
is a Project North Star Association event, we have received considerable help
with planning from the staff of the Museum. This is a good example of the
co-operative working relationship between the Association and Museum.

During the summer we expect to show off the aircraft to the public, as usual,
on Canada Day. We will also be showing the North Star during Doors Open Ottawa
at the beginning of June, if the Museum decides to take part in the event this
year.

During 2018 the Museum organised two very enjoyable events for volunteers.
Invitations were given all Museum registered volunteers including North Star
volunteers. In the summer we once again had a Bush Pilots' Breakfast at which
the Museum staff cook a great breakfast for everyone attending (there is
nothing so good as seeing the Director General, Chris Kitzan flipping
pancakes). Shortly before Christmas the Museum organised a new event, a pot
luck lunch for staff and all the Museum volunteers---another thoroughly good
overeating occasion.

One of the benefits of becoming an active member of PNSAC is for the
comradeship with other members who all have an interest in aviation. It is not
necessary to be a volunteer to enjoy the company of other members. Once a month
there is an informal breakfast held in a caf\'{e} on Arch St which is relatively
near the Ottawa General Hospital. The breakfast started about two years ago
under the leadership of Bill Tate. Each month we chat about aviation, the North
Star project and whatever else we feel like talking about. I usually give the
group an update on anything of interest that has happened since the last
breakfast. Joining this informal group is a good way to become involved with
the Association and to possibly move on to taking a more active role in the
Association.

We will be holding our Annual General Meeting at the Museum on Saturday, April
13. At this meeting we re-elect our Board of Directors and would welcome the
nomination of any member who might be interested in serving on our Board. At
present we have five members. This leaves one vacancy to bring us up to our
normal Board number of six members.

2019 is starting well. We look forward further significant milestones in the
North Star restoration process such as rehanging of the restored Engine \#4. We
are also hoping to arrange further events for our Association members.

### 2019-01-13 pnsac/newsletter/article_ProgReport.tex
January 2019 changes and additions.

\title{Conservator's Corner}
%\subtitle{PNS 2016 Status Update}
\author{R\'{e}jean Demers}

@@

\textit{R\'{e}jean Demers (Rej, to those who work with him) is a Transport Canada
licensed Aircraft Maintenance Engineer (AME). R\'{e}jean was a small
business owner in Gatineau QC for over 10 years, working for a wide
variety of private and commercial aircraft owners and operators.
Following a stint in Northern Ontario working on turpoprop freighters
and bulk fuel haulers, he took up his present position as a
Conservator, Special Projects Manager with Ingenium, Canada's Museums
of Science and Innovation in January 2017. His main project is the
restoration of the Canadair North Star DC-4GM at the Canadian Aviation
and Space Museum. He recounts below his introduction to his new job.}

Leaving Pickle Lake was like getting shot out of a cannon in a snow
storm. At that time, Christmas of 2016, I was the last person on my
crew to leave the YPL base, having seen my colleagues off for the end
of our tour and the holidays. After a full day of flying, the fleet had
to be tucked away in the barn ready for service. On Boxing Day, a fresh
crew would fly in and start another two week tour of line maintenance
with the company we worked for. However, on that Christmas, I was
quitting my job.

Nobody knew I was leaving until I got back to the bunkhouse and shared
Christmas dinner with the remaining pilots and ops personnel. I refer
to them in these terms now, but then and there, I could say they were
my friends. In the weeks leading to my departure, no one knew I was
leaving. No one speaks of breaking out of prison when they are behind
bars. The difference being the only bars I saw were the Pickle Lake
Hotel and a watering hole in Thunder Bay called the Apollo.

Driving down the 599 in that snow storm with a truck I had bought for
\$100 was a white knuckle ride. Knowing this would probably be the last
time I drove the 2000km distance between Northern Ontario and home in
Gatineau. The stakes were high: A flight to catch in Ottawa (T-minus
48hrs), New Year's eve in New York city with my family waiting there
for me and a new job starting at CASM on the day of my wedding
anniversary. Could 2017 start with anything less? Arriving to work that
January 3rd morning was a good feeling, until I got startled by "Manny"
the mannequin sitting in my predecessor's office chair. If a sense of
humour was a job requirement, I hoped to get the joke. I would later
find Manny in other compromising positions, some of which I refuse to
admit my participation in writing.

The momentum I have gained in the twelve months since my escape from
Pickle Lake is only part of what keeps me moving forward. I also count
on the support of others, namely co-workers and volunteers. My position
as Project Manager of the North Star restoration relies on the good
work and cooperation of all hands involved. As the work itself proves
to be technical and highly detailed, with wide ranges of abilities
involved, people are always the driving force.

It was suggested my article for the North Star Chronicle be focused on
the restoration of a certain four engined artifact \#1967.0645. For
details on the work that has been carried out by our dedicated group of
volunteers, I recommend you read my conservation report on the subject.
It is available online: \textcolor{blue}{\texttt{\url{http://www.projectnorthstar.ca}}}
or by email: \textcolor{blue}{\email{rdemers@techno-science.ca}}

%\section{Nr 4 Engine}
%\label{sec:engine_4}
%
%Work has continued on much of the ancillary equipment for the engine. The
%supercharger and intercooler have been completed and installed, as well as the
%reduction gearbox. The radiators and header tank have been restored and painted, and
%two of the three radiator sections have been installed. Restoration work has
%begun on the air intake system, and many of the fuel and oil lines. 
%
%\begin{figure}[htbp]
%   \vspace{2em}
%   \centering
%   \includegraphics[scale=0.5]{"PNS 007-im2".png}
%   \caption*{\small \em Volunteer Peter Trobridge inspecting the air intake fairing from engine \#4 prior to beginning restoration work.}
%   \label{fig:engine_no_4_2}
%\end{figure}
%
%\begin{figure}[htbp]
%   \vspace{2em}
%   \centering
%   \includegraphics[scale=0.5]{"PNS 003-im3".png}
%   \caption*{\small \em Volunteers Robert Desjardins, Bruce Gemmill and Michel Cote hold the refinished coolant tank from engine \#4.  Hours of polishing were required to get the tank to look like new.}
%   \label{fig:coolant_tank}
%\end{figure}
%
%The starter motor was found to be very corroded, and disassembly and repair not viable. A
%new unit has been ordered and will be installed once delivered. The fire
%suppression lines are complete and will be installed along with the cowls when
%ready. Some of the cowl panels have been restored, although it has been found
%that this engine has more corrosion than the other three engines, possibly it
%had been in service longer? This has meant that more work is required to
%refurbish each of the many panels. We still hope to have this engine ready for
%installation by next spring. 
%
%\begin{figure}[htbp]
%   \vspace{2em}
%   \centering
%   \includegraphics[scale=0.5]{"PNS 008-im1".png}
%   \caption*{\small \em Volunteer Garry Dupont working on the re-assembly of engine \#4.}
%   \label{fig:engine_no_4}
%\end{figure}
%
%\section{Fuselage and Main Cabin}
%\label{sec:main_cabin}
%
%With the aircraft moved outside in March, little work was done in the aircraft
%over the winter. Once the temperatures warmed up, we continued work in the main
%cabin to prepare for completion of the painting started last year. Repairs to
%the rear bulkhead are now complete with the installation of several new pieces
%of aluminum to replace corroded ones. A new section of flooring was installed
%where the rear washroom was situated. This required replacing about 1000 rivets!
%
%\begin{figure}[htbp]
%   \vspace{2em}
%   \centering
%   \includegraphics[scale=0.5]{"pns 043-im4".png}
%   \caption*{\small \em A single new floor panel at the rear of the main cabin required over 1000 rivets.  
%   Cleco clips are used to accurately align the panel during riveting.}
%   \label{fig:panel_revets}
%\end{figure}
%
%Cleaning and corrosion removal continued throughout the cabin. It seems whenever
%we feel we are near completion, we find more corrosion. We had also planned to
%finish priming the floor and painting the rear bulkhead before the aircraft
%returned to the hanger later in the year, but this was delayed by the shutdown
%of the project while a new Project Manager was hired by the museum. The wood
%floors and walls still need to be repaired, or new panels made, and all will
%require painting. 
%
%\section{Cabin Liners}
%\label{sec:liners}
%
%All the large pieces of liner have either been repaired or replaced, but with
%limitations on space, most remaining work on the cabin liners has been
%postponed. New window surrounds have also been made, but then all liners will
%require painting and stenciling, and this must wait until there is more room in
%the shop and the paint booth. Liners were also made for the top of the heater
%ducts.
%
%\begin{figure}[htbp]
%   \vspace{2em}
%   \centering
%   \includegraphics[scale=0.5]{P1020369-im5.png}
%   \caption*{\small \em One of the refurbished heater duct sections with a new fabric headliner secured to the top. The sides are covered with mica insulation that was removed, cleaned and reattached after repairs to the metal duct. These will be installed in the main cabin.}
%   \label{fig:heaters_liners}
%\end{figure}
%
%\section{PNS 2017 Update}
%\label{sec:pnsupdate}
%
%We were able to restart work on the aircraft in late January, following the
%arrival of our new Project manager, Rej Demers. So far, we have completed the
%top cowl panels and one of the radiator flaps, which required a repair patch to
%the outer skin, and the electrical harnesses, which are now waiting for
%painting. Work continues on the air box and the remaining cowl panels.
%
%We have decided not to move the aircraft outside this year. With our efforts
%concentrated on completing engine \#4, it is unlikely we would have the resources
%to also complete the repair work in the cabin needed to allow for priming and
%painting, which is the main reason for moving the aircraft outside in the summer.

>>>>

\title{Conservator's Corner}
%\subtitle{PNS 2016 Status Update}
\author{R\'{e}jean Demers}
\author{Restoration Project Manager}

@@

\section{Zen and the Art of Aircraft Restoration.}

Through my daily experience over the past two years, working with volunteers, I
have asked myself the following question: What makes a volunteer dedicated to a
project? The answer is not a straightforward one. This article explores the
subject, much like Pirsig did in his book about metaphysical motorcycles; Zen
and the Art of Motorcycle Maintenance.

We strive for quality. As restoration work continues on the North Star, we get
absorbed into the weight and breadth of the subject matter. This is the kind of
project that demands the level of commitment found in monasteries, dojos and
Paratrooper units. Commitment comes from dedication and reward. Quality is
obtained through application of commitment, producing results which may be
enjoyed. The following effort in writing has been made to define such elements,
those necessary to sustain the project.

A voracious appetite for detail. Precise facts and data build foundations.
Foundations of information that erode over time. Lost to floods, fires, or
worst; thieves, vandals and neglect. Information is our firmament that keeps us
grounded in our work. Daily worksheets are filled with steps taken to satisfy a
condition. Drawings are made to translate reality to paper. These words,
drawings and tables form hard to translate time lines. Coloured bars overtaking
each other, running in thousand hour blocks of volunteer contributions. These
life hours, cannot be overlooked, as a restoration project advances through the
years.

Inorganic compounds risk distraction. Focus is important when avoiding the
ethereal draw of solvents. A volunteer scrubber, washing away years of residue,
layered upon a finish that once gleamed with pride. Slowing the return to dust,
so feared and fought by the living. A process to follow and adhere to. The
methodical dismantling and reconstruction of an object. Return to form, that
which was almost indiscernible. The result of all his work is not merely a
shiny piece of aluminium panelling; He has travelled through time.

More than just an object, we look to a hidden past. Beyond these material
trappings, lay the history of Canadair North Star 17515 and her sisters.
Examples of service are easily found in books and photos. What proves a
challenge is explaining how a bumblebee becomes a preserved specimen in an
engine, how graffiti made during heavy maintenance is the only evidence of a
man's past. Where the dirt under floorboards, came from countries that no
longer bear the name she visited.

The static, prevents complete appreciation. She's more than 45,000 lbs. of
metal. Superficial understanding can only consider her 39 years an outdoor
relic. Likewise, so would a skin deep treatment of a last remaining artefact of
this type. What impact she had before the stillness of the seasons is not lost
to those who look further.

A biological experience in personal evolution. The volunteer has developed new
muscles and sinew adapted to producing polished surfaces. An eye for
microtexture and grain direction has developed. Sympathetic scars are
accumulated through careful little sacrifices. Navigating the disruptions of
capricious human needs such as rest, nourishment and leisure. Through the
application of labour, a piece of history is preserved.

Disappearing horizons. Meditative expanses of aluminum, exhibiting varieties of
corrosion rarely encountered in a lifetime's career. Cavernous interior spaces,
where cargo of all sorts was bound for ports beyond. Flight control surfaces
that have pressed through all shades of dusk, emerging in the dawn of a new
day. Now removed from the ship, these weathered, beaten, bruised remnants of a
glorious past await restoration. A new horizon forms, some many years away,
known to those who are on-board for the long run.

Healthy obsessions. Volunteer restoration technicians take home an experience.
They take to heart the results of their labour, in forming a product that is
also intangible. The mind has strengthened, forming resolves that assist,
understand and process the elements of restoration work. As historians,
fanatics and unwitting library wanderers soon discover; The story of air
transport, in its beginning, is far from just a few obscure paragraphs in
Canada's ongoing history. It lives through those people who continue writing
it, preserving the physical, extending beyond.
